```{r inline_args, echo=FALSE}
args <- commandArgs(TRUE)

query = args[1]
inFile= args[2]
#query = "mouse liver Chip-seq embryo Illumina"
#inFile="/home/karo/Documents/tests/cat_retina_rna-seq_Table.tab"
output_path=dirname(normalizePath(inFile))
```
---
title: "Metadata summary"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# make directory for figures
subDir ="figs_tmp"
dir.create(file.path(output_path, subDir), showWarnings = F)
figs_tmp_dir=file.path(output_path, subDir)

```

## This is a summary for selected metadata obtained from the SRA repository with the query search : "`r query`".

```{r libs and functions, include=FALSE}
# import packages
library(knitr)
library(ggplot2)
library(tibble)
#library(kableExtra)
library(dplyr)

###############################  Functions ################################

# make barplot for input metric
metric_barplot <- function(df, metric) {
  ggplot(data= df, 
         aes(x = .data[[metric]], 
             fill = .data[[metric]])) +
    theme_classic() +
    ggtitle(paste0("Bar chart of ", metric, "\n"))+
    theme(plot.title = element_text(size=13))   +
    theme(legend.text=element_text(size=9))+
    theme(legend.title=element_text(size=12))+
    theme(legend.position = "bottom")+
    guides(fill = guide_legend(ncol = 3,title.position = "top",title.hjust = 0.5))+
    theme(axis.title.y=element_text(size=11))+
    theme(axis.ticks.x = element_blank())+
    theme(axis.title.x=element_blank())+
    theme(axis.text.x = element_blank()) +
    ylab("Number of samples\n") +
    geom_bar()+
    geom_text(aes(label = after_stat(count)),stat="count", vjust = -0.2, size=3)

}

# make png plot
make_png <- function(plottt, title){
  png(title, width = 960, height = 960,res = 150)
  print(plottt)
  cat("\n")
  cat("\n")
  invisible(dev.off())
}

## split metadata into list of data_frames
split_meta_data_on_study_accession <- function(df,n, n_samp){
  # """ splits meta data into several data frames each including 10x meta data for a max of "n" samples. 
	#Take as input data frame (df) and number of samples per data_frame in the list (n) plus the number of unique accession ids in df. """
  #print(n_samp)
  df_list <- list()
   if (n_samp <= n) {
      df_list[[1]] <- df
   } else {
   # make a list of tables
      # number of DF to split into so that there is max samples_per_table samples per df
      n_df <- ceiling(n_samp/n)#round(n_samp/n,digits = 0)
     # print(n_df)
      # set first row number
      first_row = 1
      # iterate over data_frames number to add elements of the list
      for (i in c(1:(n_df-1))) {
        #print(i)
         last_sample_name=unique(df$study_accession)[i*n]
         last_row = max(which(df$study_accession==last_sample_name))
         df_list[[i]] <- df[first_row:last_row,]
         first_row = last_row+1
      }
      # last data frame in the list
      previous_sample_name=unique(df$study_accession)[(n_df-1)*n]
      f_row = max(which(df$study_accession==previous_sample_name))+1
      df_list[[n_df]] <- df[f_row:nrow(df),]
      }
   df_list
}

```

```{r get data, include=FALSE}
## read data 
mydf <- read.delim(inFile)

# show number of samples
ns <- nrow(mydf)
cn <- matrix(colnames(mydf), length(colnames(mydf)))
# show number of study_accession:
sa=unique(mydf$study_accession)
sal=length(sa)

# select relevant metrics to show in barplot
mm=c("organism_name", "library_source","library_selection",
     "instrument", "library_layout","instrument_model_desc")
## add metrics for dot plots
mmall=c(mm,"run_total_spots","run_total_bases")
# conv to matrix for print
mmp <-matrix(mmall, length(mmall))

cat(" ")

```
# Total samples retrieved : `r ns`

# Different study_accession ids : `r sal`

# Metadata available : 

```{r meraics, echo = F}
# print colnames
kable(cn, "simple")

cat(" ")

```
# Metadata selected for graphics :

```{r selected, echo = F}
kable(mmp, "simple")

cat(" ")

```
# Bar charts for selected metadata



## Description of selected metadata

**instrument** - type of instrument used for library sequencing. Note differences in chemistry characteristics (list non exhaustive):

	- Illumina Genome Analyzer, Hiseq 2000, 2500 - SBS four color chemistry, random flow cell

	- Illumina Hiseq 4000, Hiseq X Ten - SBS four color chemistry, patterned flow cell

	- Illumina Novaseq 6000 - SBS two color chemistry, patterned flow cell 

	- Illumina Next Seq 500 and 550 - SBS two color chemistry, random flow cell

	- PacBio RS II, Sequel, Sequel II - long read sequencing with SMRT technology
	
	- Oxford Nanopore, MinION, PromethION - native long read sequencing 

**instrument_model_desc** - note presence of differences in sequencing technology used by Illumina, PacBio, Oxford Nanopore, Ion Torrent (list non exhaustive)

**library_layout** 

	- PAIRED : read 1 and read 2 in forward and reverse orientation, respectively

	- SINGLE : read 1 only

**library_selection** - molecular biology method used for selecting nucleic acid of interest 

**library_source** - library derived from sample type : DNA (genomic) or RNA (transcriptomic) AMPLICON (PCR amplification of a specific genomic locus); (list non exhaustive)

**run_total_spots** - equivalent to spots on an array or sequence reads in a NGS experiment

**run_total_bases** - total number of bases in the library (function of number of reads and read length)



## Bar plot figure legend

Different metadata characteristics are represented with bars, each color corresponding to a different category.

Y axis shows the number of samples within each category. Numbers above bars show numbers of samples in a given category.

```{r plots, echo=FALSE, out.height=".9\\paperheight"}

# make plots for all selected metrics
plotss=list()
for (i in mm){
  pl <-  metric_barplot(mydf, i)
  plotss[[i]] <- pl
} 

# make file names for all selected metrics
fileNames=list()
for (i in mm){
  fn <-  paste0(figs_tmp_dir, "/",i,"_barplot.png")
  fileNames[[i]] <- fn
} 

# print PNG files
n_plotss <-length(plotss)
for (v in 1:n_plotss ) {
 make_png(plotss[[v]], fileNames[[v]])
}

# list PNG files
plots_to_include_list <- list.files(path = figs_tmp_dir, pattern = "_barplot.png" )
plots_to_include_l <- paste(figs_tmp_dir, plots_to_include_list, sep = "/")

# include in file
knitr::include_graphics(plots_to_include_l)
cat("\n")
cat(" ")

```

```{r plots dot make, echo=FALSE}

############## run_total_bases

# how many differnt accession numbers to group per study
n_samp<-length(unique(mydf$study_accession)) 
# split the meta data dataframe into a list of DF
meta_d_list<-split_meta_data_on_study_accession(df=mydf,n=40,  n_samp = n_samp)
# print run_total_spots to file
n_tab <- length(meta_d_list)
for ( i in 1:n_tab ) {
     # specify output file
     file=paste0(figs_tmp_dir, "/",i,"_spots.png")
     # make plot   
	div_factor=1e6
      crp<-ggplot(meta_d_list[[i]], aes(x=study_accession, y=run_total_spots/div_factor)) + 
              theme_classic()+
              geom_point(col="tomato2", size=3) +   # Draw points
              geom_segment(aes(x=study_accession, 
                               xend=study_accession, 
                               y=min(run_total_spots/div_factor), 
                               yend=max(run_total_spots/div_factor)), 
                           linetype="dashed", 
                           size=0.1, col="grey90") +   # Draw dashed lines
              labs(title="Dot Plot of run_total_spots in millions per library and study accession_number", 
                   subtitle="run_total_spots = number of reads")+ 
              ylab("Run total spots (millions)") +  
	      theme(axis.title.y=element_blank())+
              coord_flip()
     #print plot to file
     png(file, width =960 ,height =1300, res=150)
        suppressWarnings(print(crp))
     invisible(dev.off())
          cat( "\n" )
	  cat( "\n" )
}
## create a list with files to include in the PDF
spots_file_list<-list.files(path = figs_tmp_dir, pattern = "*spots.png")
spots_file_l <- paste(figs_tmp_dir, spots_file_list, sep = "/")



############## run_total_bases

# show total bases in a run with dotplot proportional to number of libraries per study
# calc mean bases per project
df1=aggregate(mydf$run_total_bases, list(mydf$study_accession), FUN=mean)
colnames(df1)<-c("study_accession","mean_run_total_bases")
rownames(df1)<-df1$study_accession
library(dplyr)
# calc number of libraries per study
df2 <- mydf %>% count(study_accession,sort = T)
colnames(df2)<-c("study_accession","number_of_libraries")
rownames(df2)<-df2$study_accession
# merge
df3 <- data.frame(df1, number_of_libraries=df2[rownames(df1),"number_of_libraries"])
#colnames(df3)<-c("study_accession","number_of_libraries","mean_run_total_bases")
#sort
df3<-df3[order(df3$number_of_libraries, decreasing = T),]
# split
meta_d2_list<-split_meta_data_on_study_accession(df=df3, n=40, n_samp = n_samp)
# plot
n_tab <- length(meta_d2_list)
for ( i in 1:n_tab ) {
     # specify output file
     file=paste0(figs_tmp_dir, "/",i,"_bases.png")
     # make plot    
     div_factor=1e6
     dp <-  ggplot(meta_d2_list[[i]], aes(x=study_accession, y=mean_run_total_bases/div_factor, label=number_of_libraries)) + 
          theme_classic()+
          geom_point(col="lightblue", aes(size=number_of_libraries), stat='identity' )+   # Draw points
          geom_text(color="black", size=2, nudge_x = 0.3)+
          geom_segment(aes(x=study_accession, 
                           xend=study_accession, 
                           y=min(mean_run_total_bases/div_factor), 
                           yend=max(mean_run_total_bases/div_factor)), 
                       linetype="dashed", 
                       size=0.1, col="grey90") +   # Draw dashed lines
          labs(title="Dot Plot of mean run_total_bases in millions  per study_accession", 
               subtitle="Dot size corresponds to number of libraries per study_accession")+ 
          ylab("Mean run total bases (millions)") + 
	  theme(axis.title.y=element_blank())+
	  theme(legend.position = "bottom")+
	  theme(legend.direction = "horizontal")+
	  coord_flip()
    #print plot to file
     png(file, width =960 ,height =1300, res=150)
        suppressWarnings(print(dp))
     invisible(dev.off())
          cat( "\n" )
          cat( "\n" )
}

## create a list with files to include in the PDF
base_file_list<-list.files(path = figs_tmp_dir, pattern = "*bases.png")
base_file_l <- paste(figs_tmp_dir, base_file_list, sep = "/")



```
\newpage

# Dot plots for selected metadata give an appreciation of the number of libraires per study and data volume



##  Dot plots for run_total_spots per library in study_accession

X axis shows run_total_spots in millions.

Study accession number is on the vertical axis. Each dot shows a number of single-end reads (run_total_spots) per library 
for a given study_accession id. 


##  Dot plots for mean run_total_bases per library in study_accession

X axis shows mean value of run_total_bases in millions.

Study accession number is on the vertical axis. Each dot represents mean run_total_bases in a given study_accession id. Dot size is proportional to the total number of reads (single and paired) in a study.

```{r plots print, echo=FALSE}

knitr::include_graphics(spots_file_l)

knitr::include_graphics(base_file_l)

```

```{r end,include=F}


